---
title: "Behaviour and Onboard Video"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Behaviour and Onboard Video}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

There are two ways to know what a tagged animal was doing. The **sensors** tell you how it
moved -- how hard it beat its tail, how it turned, whether it was in day or night. The
**camera**, if the tag carried one, tells you what actually happened -- a feeding lunge, a
social interaction, a cleaning station. This guide covers both, and the point where they
meet: joining what the camera *saw* to what the sensors *felt* gives you labelled data, and
labelled data is what turns a biologging dataset into a behavioural classifier.

Both branches start from a processed `nautilus_tag`. If you do not have one yet, see
`vignette("getting-started", package = "nautilus")`. As before, the chunks below are
illustrative -- they show the real calls in order but are not executed here.

# Part 1 -- Behaviour from the sensors

## Diel phase

Almost every behavioural question benefits from knowing whether a sample fell in day, night
or twilight. `getDielPhase()` computes that from the timestamp and the deployment location
using the NOAA solar-geometry equations (no geospatial system libraries required).

```{r diel}
library(nautilus)

tag1  <- tags[[1]]                 # a single deployment from the processed list
meta1 <- getTagMetadata(tag1)
coords <- matrix(c(meta1$deployment$lon, meta1$deployment$lat), ncol = 2)  # lon, lat

tag1$diel <- getDielPhase(tag1$datetime, coords, phases = 3, solar.depth = 12)
```

`phases` chooses the resolution: `2` gives `day`/`night`, `3` adds a merged `crepuscule`, and
`4` splits it into `dawn`/`day`/`dusk`/`night`. `solar.depth` sets the twilight definition --
`6` (civil), `12` (nautical) or `18` (astronomical). It returns a plain factor the same
length as the input, so you assign it straight onto a column. (Note the coordinate order is
**longitude first, then latitude**.)

## Tail-beat frequency

How hard and how often an animal beats its tail is a direct readout of swimming effort and
gait. `calculateTailBeats()` reads that rhythm straight from a band-passed acceleration
channel.

```{r tailbeats}
tags <- calculateTailBeats(
  tags,
  method      = "peaks",   # per-beat detection (frequency + amplitude)
  motion.col  = "sway",    # lateral axis for a laterally-swimming shark
  min.freq.Hz = 0.1,
  max.freq.Hz = 2.5        # sampling rate must exceed 2x this; >= 4x is comfortable
)
```

Each backend names its own output, so a value's provenance always travels with it: `"peaks"`
adds `tbf_hz_peaks` and `tbf_amplitude_peaks`, and `"wavelet"` adds `tbf_hz_wavelet` and
`tbf_amplitude_wavelet`. There is no method-agnostic `tbf_hz`. Both amplitudes are the
peak-to-trough excursion in the units of the motion axis, so the two are directly comparable.
Use `tailBeatColumn()` when you want backend-agnostic code:

```{r tbf-resolve, eval = FALSE}
col <- tailBeatColumn(tags[[1]])        # "tbf_hz_peaks" by default
mean(tags[[1]][[col]], na.rm = TRUE)
```

The function also adds `tbf_swimming`, which is *not* suffixed -- it comes from the band-passed
signal both backends share, so it is the same whichever ran. It is `NA` unless you supply a
`min.amplitude` reference:
whether an oscillation is active swimming rather than gliding (or, on a towed tag, flow and
tether motion) cannot be told reliably from the sway signal alone, so the function reports the
frequency and amplitude and leaves the behavioural call to you. The two methods fail in
opposite directions -- peak detection can lock onto a harmonic and report a multiple of the
true frequency, while the wavelet can be captured by a
strong low-frequency component that peak detection ignores -- so where they agree, they are
worth more than either alone.

Two things govern the choice of axis and rate. Pick the axis the animal actually beats on:
`"sway"` (lateral) for most sharks and teleosts, `"heave"` for animals that beat vertically
(cetaceans, some rays) -- feeding `"surge"` or `"heave"` for a laterally-swimming fish can
report double the true frequency, and the function warns when it detects this. And respect
the Nyquist limit: you need at least twice `max.freq.Hz` to resolve the beat (the run aborts
below that), and four times to be comfortable.

## Candidate circling events

`detectCircling()` identifies sustained, predominantly same-direction heading rotations and returns
one row per event. It is a specialised query, not a new structuring step: dive detection and track
reconstruction are optional, and the original sensor records are unchanged.

```{r circling}
circles <- detectCircling(
  tags,
  min.rotations = 2,
  min.directionality = 0.8,
  variables = c("depth", "temp", "vedba")
)

# The same event table is accepted by both plotters. No sample annotation is required.
plotDepthProfiles(tags, events = circles)
tracks <- reconstructTrack(tags)
plotTracks(tracks, events = circles)

# Optional sample labels: 1 inside events, 0 elsewhere when assessed, NA when unassessable.
labelled <- annotateData(tags, circles)
attr(circles, "circling_detection")$deployments
saveRDS(circles, "circling_events.rds")   # preserves settings and assessment windows
```

The defaults are heuristic starting points. Heading must change by less than 180 degrees between
successive observations; gaps, missing orientation and near-vertical posture break assessment.
Smoothing and pause tolerances are in seconds, and angular rates are in degrees/second. Known pitch
offsets are accounted for in the posture screen. Very slow circling may require a smaller
`min.turn.rate`; fast-turning animals may require finer sampling and shorter smoothing.

To inspect one dive, subset a deployment to its `dive_id` and pass the original event table to
`plotDepthProfiles(events = circles)`. The plot clips windows without changing the event metrics.
Events are detected over the full deployment and may cross dive or phase boundaries. A missing band
does not demonstrate absence if heading was unassessable, and loops in a reconstructed track are
not independent confirmation: both calculations use heading. Infer ecological function from
additional depth, activity or video evidence.

### Matching events to dive intervals

For custom analyses, an overlap join preserves all event--dive relationships rather than forcing
each event into one dive. The following recipe makes a separate relationship table:

```{r circling-dive-overlap}
dives <- detectDives(tags)
dm <- diveMetrics(dives)
ev <- data.table::as.data.table(circles)[, .(ID, circling_id, event_start = start, event_end = end)]
di <- data.table::as.data.table(dm)[, .(ID, dive_id, dive_start = start, dive_end = end)]
data.table::setkeyv(di, c("ID", "dive_start", "dive_end"))
links <- data.table::foverlaps(
  ev, di,
  by.x = c("ID", "event_start", "event_end"),
  by.y = c("ID", "dive_start", "dive_end"),
  type = "any", nomatch = NULL
)
links[, overlap_s := as.numeric(
  difftime(pmin(event_end, dive_end), pmax(event_start, dive_start), units = "secs")
)]
links <- links[overlap_s > 0]  # endpoint contact alone contributes no elapsed time
```

For phase-specific analysis, derive intervals from contiguous valid runs of `dive_phase` rather
than treating a gap-spanning first/last timestamp as uninterrupted bottom residence. Do not copy an
event's total rotations to every matching dive, or prorate them by duration without an explicit
constant-rate assumption. Union overlapping windows for time budgets and use assessed duration as
the denominator. A mean binary annotation is a sample fraction, not an automatic elapsed-time budget
under irregular sampling.

## Windowed features for machine learning

To classify behaviour you rarely feed raw samples to a model; you summarise short windows.
`extractFeatures()` computes per-window statistics for any set of variables over a sliding or
non-overlapping window.

```{r features}
feats <- extractFeatures(
  tags,
  variables   = c("vedba", "pitch", "odba"),
  metrics     = c("mean", "sd", "max"),
  window.size = 5            # seconds
)
```

You must supply both `variables` and `metrics` (unless you pass a `parameter.grid` for
per-feature window sizes). Beyond the linear statistics (`mean`, `sd`, `energy`, `entropy`,
...), `enhanced.features = TRUE` unlocks a family of movement metrics -- heading change,
circling, posture stability, activity index -- computed with proper circular statistics
(needs the **circular** and **zoo** packages). Circular variables (by default `heading` and
`roll`) are handled on the circle rather than as plain numbers. We return to the
`response.col` argument -- the one that attaches a behavioural label -- in Part 3.

# Part 2 -- Onboard video

## Recovering each clip's timestamp

Everything downstream depends on knowing exactly when each video frame was recorded, in
sensor time. `getVideoMetadata()` recovers that **filename-first**: it reads the start time
from the video's file name wherever the camera encodes one (exact, and independent of any
on-screen clock), and falls back to reading the burned-in overlay clock by OCR only where the
name carries no time.

```{r video-metadata}
clock_corrections <- getVideoClockCorrections(tags)
clock_corrections                         # review before applying

vmeta <- getVideoMetadata(
  "./videos",
  timestamp.source = "auto",   # file name first, OCR only where needed
  cross.check      = TRUE,     # also OCR the overlay and flag disagreements > 2 s
  clock.corrections = clock_corrections
)
```

The result is one row per clip with `start`, `end`, `duration`, `frame_rate` and a
`timestamp_source` column; `cross.check = TRUE` adds an `ocr_flag` that marks any clip whose
file-name time and on-screen time disagree. Feed the result to
`saveUncertainTimestampFrames()` to dump the timestamp crops for every clip that is
OCR-sourced, missing, or flagged, so you can confirm them by eye:

```{r review-frames}
vmeta <- saveUncertainTimestampFrames(vmeta, output.dir = "./timestamp-review")
```

If the sensor import retained a CATS sidecar whose `[logging]` clock is explicitly UTC, the sensor object
was imported in UTC, and the `[device]` clock has a non-zero offset, `getVideoClockCorrections()` puts the
corresponding correction in the reviewable table above. No correction is inferred from ambiguous or
non-UTC declarations.

For a correction established independently, supply the same compact contract manually. The value is
the number of seconds **added** to the extracted video clock:

```{r override}
clock_corrections <- data.frame(
  ID = "PIN_CAM_31",
  clock_correction_s = 3600,
  clock_correction_source = "manual"
)

vmeta <- getVideoMetadata("./videos/PIN_CAM_31", clock.corrections = clock_corrections)
```

Every returned row records `clock_correction_s` (`0` when uncorrected) and
`clock_correction_source` (`NA` when uncorrected), so the original reading remains recoverable as
`start - clock_correction_s` without duplicating the timestamp columns.

**Dependencies and the OCR model.** `ffprobe` (from FFmpeg) is always needed for duration and
frame rate; a pure filename run needs nothing more. OCR additionally uses `ffmpeg` and the
**tesseract** and **magick** packages, plus a fine-tuned camera-tag OCR model (~11 MB) that is
downloaded on first use and cached -- run `installCamOcrModel()` to pre-fetch it for offline
work, and it falls back to Tesseract's generic model if unobtainable. The OCR settings (the
on-screen-clock box geometry, search depth) live in `ocrControl()`; the defaults target a 4K
camera's bottom-right overlay, so a different camera needs a tuned
`ocr = ocrControl(box = ..., frame.height = ...)`.

## Re-encoding for archival (optional)

Raw camera-tag footage is bulky. `reencodeVideos()` batch-transcodes a folder to
space-efficient HEVC (H.265) via FFmpeg -- useful before archiving or shipping clips.

```{r reencode}
reencodeVideos("./videos/raw", output.dir = "./videos/hevc")
```

It picks a hardware encoder when one is available and strips audio for size. (If you write
back to the same folder, set a `file.suffix` so an `.mp4` input is not skipped for having the
same output path.)

## Jumping to a moment, and overlaying the sensors

To inspect a specific instant -- say a manoeuvre surfaced by `findValidationSegments()` (see
`vignette("movement-tracks")`) -- `launchVideo()` opens the covering clip in VLC, seeked to
that exact time:

```{r launch}
launchVideo("PIN_CAM_01", as.POSIXct("2023-08-31 17:40:00", tz = "UTC"), vmeta)
```

And to see the sensors and the footage together, `renderOverlayVideo()` composites a live,
time-synchronised sensor dashboard beside the video. The default orientation module combines a
three-dimensional attitude model, a heading compass, and numeric heading/pitch/roll values:

```{r overlay}
renderOverlayVideo(
  vmeta$file[1], tag1, "./overlay/PIN_CAM_01.mp4",
  dashboard   = "general",       # 3-D orientation + depth, VeDBA and vertical velocity
  video.start = vmeta$start[1]   # the sync anchor, from getVideoMetadata
)
```

For a less intrusive result, put selected modules directly over the source footage. This keeps the
source dimensions and uses local translucent backplates rather than a full opaque panel:

```{r overlay-custom}
renderOverlayVideo(
  vmeta$file[1], tag1, "./overlay/PIN_CAM_01_compact.mp4",
  dashboard   = "compact",
  metrics     = c("orientation", "depth", "vedba"),
  composition = "overlay",
  side        = "right",
  video.start = vmeta$start[1]
)
```

Set `orientation = "dials"` to replace the combined model with separate heading, pitch, and roll
dials. When the combined model is used, `orientation.model` selects a schematic `"shark"`,
`"cetacean"`, `"turtle"`, `"fish"`, or `"manta"` silhouette without changing the sensor values or
body-axis convention. The cetacean is dolphin-like, the turtle is a hard-shelled sea turtle, and the
fish is a generic ray-finned fish; none of the models is a species-specific reconstruction.
The `"ribbon"` preset places time-series modules across the lower third, while `"focus"`
enlarges the first selected metric. After `reconstructTrack()`, `dashboard = "expanded"` adds a
fixed-camera 3-D pseudo-trajectory whose past path, future context, surface projection, metric scale,
and vertical exaggeration are explicit; gaps in the sensor record are not joined by false straight-line
segments.

The fixed `"validation"` dashboard draws a large 3-D attitude indicator with heading deliberately
omitted (roll, not heading, is the cue for a handedness check); `"validation-compare"` shows competing
mappings side by side. `renderOverlayVideo()` needs FFmpeg on the PATH and the **av** package, and takes
a **single** deployment's data -- pass one element (`tags[[1]]`), not the whole list. At normal and
detailed verbosity, dashboard rendering and FFmpeg encoding have separate progress indicators.

# Part 3 -- Where the two branches meet

Now the payoff. You have behavioural events scored from the footage -- a data frame of
`start`/`end` windows per animal, each tagged with an event type. `annotateData()` joins those
onto the sensor data, adding a binary column (`1` inside the window, `0` outside) for each
event type:

```{r annotate}
tags <- annotateData(
  tags,
  annotations     = annotations,   # id / event / start / end, from your video review
  event.col       = "category",
  selected.events = c("feeding")
)
```

Because unreviewed stretches (night, poor visibility) carry no annotations, a raw `0` there
would be a false negative. `filterVideoPeriod()` restricts the sensor data to the intervals
that were actually filmed -- and, optionally, to the intervals you actually reviewed -- so the
labels mean what you think they mean:

```{r filter-period}
tags <- filterVideoPeriod(
  tags,
  video.metadata       = vmeta,
  annotation.intervals = annotation_intervals   # optional: only the reviewed windows
)
```

With clean labels on the sensor stream, `extractFeatures()` closes the loop: point
`response.col` at the label column and it carries the label through to each window (aggregated
by majority vote), producing a feature table ready to train a classifier.

```{r labelled-features}
model_data <- extractFeatures(
  tags,
  variables    = c("vedba", "pitch", "roll"),
  metrics      = c("mean", "sd"),
  window.size  = 5,
  response.col = "feeding"       # the label added by annotateData()
)
```

From here you are in ordinary supervised-learning territory -- a random forest, say -- with
each row a window of sensor features and a behaviour label validated against video.

## Where to go next

- **Getting Started** -- the pipeline that produces the processed tag:
  `vignette("getting-started", package = "nautilus")`.
- **Movement tracks** -- the fourth branch, and the source of `findValidationSegments()`:
  `vignette("movement-tracks", package = "nautilus")`.
- **Function reference** -- `?calculateTailBeats`, `?extractFeatures`, `?getVideoMetadata`,
  `?annotateData`, `?renderOverlayVideo`, `?ocrControl`.
- **Questions and bugs** -- <https://github.com/miguelgandra/nautilus/issues>.
