---
title: "Behaviour and Onboard Video"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Behaviour and Onboard Video}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

There are two ways to know what a tagged animal was doing. The **sensors** tell you how it
moved -- how hard it beat its tail, how it turned, whether it was in day or night. The
**camera**, if the tag carried one, tells you what actually happened -- a feeding lunge, a
social interaction, a cleaning station. This guide covers both, and the point where they
meet: joining what the camera *saw* to what the sensors *felt* gives you labelled data, and
labelled data is what turns a biologging dataset into a behavioural classifier.

Both branches start from a processed `nautilus_tag`. If you do not have one yet, see
`vignette("getting-started", package = "nautilus")`. As before, the chunks below are
illustrative -- they show the real calls in order but are not executed here.

# Part 1 -- Behaviour from the sensors

## Diel phase

Almost every behavioural question benefits from knowing whether a sample fell in day, night
or twilight. `getDielPhase()` computes that from the timestamp and the deployment location
using the NOAA solar-geometry equations (no geospatial system libraries required).

```{r diel}
library(nautilus)

tag1  <- tags[[1]]                 # a single deployment from the processed list
meta1 <- tagMetadata(tag1)
coords <- matrix(c(meta1$deployment$lon, meta1$deployment$lat), ncol = 2)  # lon, lat

tag1$diel <- getDielPhase(tag1$datetime, coords, phases = 3, solar.depth = 12)
```

`phases` chooses the resolution: `2` gives `day`/`night`, `3` adds a merged `crepuscule`, and
`4` splits it into `dawn`/`day`/`dusk`/`night`. `solar.depth` sets the twilight definition --
`6` (civil), `12` (nautical) or `18` (astronomical). It returns a plain factor the same
length as the input, so you assign it straight onto a column. (Note the coordinate order is
**longitude first, then latitude**.)

## Tail-beat frequency

How hard and how often an animal beats its tail is a direct readout of swimming effort and
gait. `calculateTailBeats()` reads that rhythm straight from a band-passed acceleration
channel.

```{r tailbeats}
tags <- calculateTailBeats(
  tags,
  method      = "peaks",   # per-beat detection (frequency + amplitude)
  motion.col  = "sway",    # lateral axis for a laterally-swimming shark
  min.freq.Hz = 0.1,
  max.freq.Hz = 2.5        # sampling rate must exceed 2x this; >= 4x is comfortable
)
```

The `"peaks"` method adds `tbf_hz` (per-beat frequency) and `tbf_amplitude` (a beat-effort
proxy, in the units of the motion axis); `method = "wavelet"` tracks the strongest in-band
oscillation through a Morlet wavelet transform and returns the same two on the same footing.
It also adds `tbf_swimming`, but this is `NA` unless you supply a `min.amplitude` reference:
whether an oscillation is active swimming rather than gliding (or, on a towed tag, flow and
tether motion) cannot be told reliably from the sway signal alone, so the function reports the
frequency and amplitude and leaves the behavioural call to you. The two methods fail in
opposite directions -- peak detection can lock onto a harmonic and report a multiple of the
true frequency, while the wavelet can be captured by a
strong low-frequency component that peak detection ignores -- so where they agree, they are
worth more than either alone.

Two things govern the choice of axis and rate. Pick the axis the animal actually beats on:
`"sway"` (lateral) for most sharks and teleosts, `"heave"` for animals that beat vertically
(cetaceans, some rays) -- feeding `"surge"` or `"heave"` for a laterally-swimming fish can
report double the true frequency, and the function warns when it detects this. And respect
the Nyquist limit: you need at least twice `max.freq.Hz` to resolve the beat (the run aborts
below that), and four times to be comfortable.

## Windowed features for machine learning

To classify behaviour you rarely feed raw samples to a model; you summarise short windows.
`extractFeatures()` computes per-window statistics for any set of variables over a sliding or
non-overlapping window.

```{r features}
feats <- extractFeatures(
  tags,
  variables   = c("vedba", "pitch", "odba"),
  metrics     = c("mean", "sd", "max"),
  window.size = 5            # seconds
)
```

You must supply both `variables` and `metrics` (unless you pass a `parameter.grid` for
per-feature window sizes). Beyond the linear statistics (`mean`, `sd`, `energy`, `entropy`,
...), `enhanced.features = TRUE` unlocks a family of movement metrics -- heading change,
circling, posture stability, activity index -- computed with proper circular statistics
(needs the **circular** and **zoo** packages). Circular variables (by default `heading` and
`roll`) are handled on the circle rather than as plain numbers. We return to the
`response.col` argument -- the one that attaches a behavioural label -- in Part 3.

# Part 2 -- Onboard video

## Recovering each clip's timestamp

Everything downstream depends on knowing exactly when each video frame was recorded, in
sensor time. `getVideoMetadata()` recovers that **filename-first**: it reads the start time
from the video's file name wherever the camera encodes one (exact, and independent of any
on-screen clock), and falls back to reading the burned-in overlay clock by OCR only where the
name carries no time.

```{r video-metadata}
vmeta <- getVideoMetadata(
  "./videos",
  timestamp.source = "auto",   # file name first, OCR only where needed
  cross.check      = TRUE      # also OCR the overlay and flag disagreements > 2 s
)
```

The result is one row per clip with `start`, `end`, `duration`, `frame_rate` and a
`timestamp_source` column; `cross.check = TRUE` adds an `ocr_flag` that marks any clip whose
file-name time and on-screen time disagree. Feed the result to
`saveUncertainTimestampFrames()` to dump the timestamp crops for every clip that is
OCR-sourced, missing, or flagged, so you can confirm them by eye:

```{r review-frames}
vmeta <- saveUncertainTimestampFrames(vmeta, output.dir = "./timestamp-review")
```

Then correct only the few clips that are genuinely wrong -- typically a camera with a mis-set
clock -- by overwriting `start` (recompute `end` from `duration`):

```{r override}
i <- which(vmeta$video == "230831-161949_CAM0bc99448_30.mp4")
vmeta$start[i] <- as.POSIXct("2023-08-31 16:19:49", tz = "UTC")
vmeta$end[i]   <- vmeta$start[i] + vmeta$duration[i]
```

**Dependencies and the OCR model.** `ffprobe` (from FFmpeg) is always needed for duration and
frame rate; a pure filename run needs nothing more. OCR additionally uses `ffmpeg` and the
**tesseract** and **magick** packages, plus a fine-tuned camera-tag OCR model (~11 MB) that is
downloaded on first use and cached -- run `installCamOcrModel()` to pre-fetch it for offline
work, and it falls back to Tesseract's generic model if unobtainable. The OCR settings (the
on-screen-clock box geometry, search depth) live in `ocrControl()`; the defaults target a 4K
camera's bottom-right overlay, so a different camera needs a tuned
`ocr = ocrControl(box = ..., frame.height = ...)`.

## Re-encoding for archival (optional)

Raw camera-tag footage is bulky. `reencodeVideos()` batch-transcodes a folder to
space-efficient HEVC (H.265) via FFmpeg -- useful before archiving or shipping clips.

```{r reencode}
reencodeVideos("./videos/raw", output.dir = "./videos/hevc")
```

It picks a hardware encoder when one is available and strips audio for size. (If you write
back to the same folder, set a `file.suffix` so an `.mp4` input is not skipped for having the
same output path.)

## Jumping to a moment, and overlaying the sensors

To inspect a specific instant -- say a manoeuvre surfaced by `findValidationSegments()` (see
`vignette("movement-tracks")`) -- `launchVideo()` opens the covering clip in VLC, seeked to
that exact time:

```{r launch}
launchVideo("PIN_CAM_01", as.POSIXct("2023-08-31 17:40:00", tz = "UTC"), vmeta)
```

And to see the sensors and the footage together, `renderOverlayVideo()` composites a live,
time-synchronised sensor dashboard beside the video:

```{r overlay}
renderOverlayVideo(
  vmeta$file[1], tag1, "./overlay/PIN_CAM_01.mp4",
  dashboard   = "general",       # heading/pitch/roll dials + scrolling depth/VeDBA traces
  video.start = vmeta$start[1]   # the sync anchor, from getVideoMetadata
)
```

The `"general"` dashboard is for presentation; `"validation"` draws a large 3-D attitude
indicator (heading deliberately omitted, since roll -- not heading -- is the cue for a
handedness check) to judge an axis mapping by eye; and `"validation-compare"` shows competing
mappings side by side. `renderOverlayVideo()` needs FFmpeg on the PATH and the **av** package,
and takes a **single** deployment's data -- pass one element (`tags[[1]]`), not the whole
list.

# Part 3 -- Where the two branches meet

Now the payoff. You have behavioural events scored from the footage -- a data frame of
`start`/`end` windows per animal, each tagged with an event type. `annotateData()` joins those
onto the sensor data, adding a binary column (`1` inside the window, `0` outside) for each
event type:

```{r annotate}
tags <- annotateData(
  tags,
  annotations     = annotations,   # id / event / start / end, from your video review
  event.col       = "category",
  selected.events = c("feeding")
)
```

Because unreviewed stretches (night, poor visibility) carry no annotations, a raw `0` there
would be a false negative. `filterVideoPeriod()` restricts the sensor data to the intervals
that were actually filmed -- and, optionally, to the intervals you actually reviewed -- so the
labels mean what you think they mean:

```{r filter-period}
tags <- filterVideoPeriod(
  tags,
  video.metadata       = vmeta,
  annotation.intervals = annotation_intervals   # optional: only the reviewed windows
)
```

With clean labels on the sensor stream, `extractFeatures()` closes the loop: point
`response.col` at the label column and it carries the label through to each window (aggregated
by majority vote), producing a feature table ready to train a classifier.

```{r labelled-features}
model_data <- extractFeatures(
  tags,
  variables    = c("vedba", "pitch", "roll"),
  metrics      = c("mean", "sd"),
  window.size  = 5,
  response.col = "feeding"       # the label added by annotateData()
)
```

From here you are in ordinary supervised-learning territory -- a random forest, say -- with
each row a window of sensor features and a behaviour label validated against video.

## Where to go next

- **Getting Started** -- the pipeline that produces the processed tag:
  `vignette("getting-started", package = "nautilus")`.
- **Movement tracks** -- the fourth branch, and the source of `findValidationSegments()`:
  `vignette("movement-tracks", package = "nautilus")`.
- **Function reference** -- `?calculateTailBeats`, `?extractFeatures`, `?getVideoMetadata`,
  `?annotateData`, `?renderOverlayVideo`, `?ocrControl`.
- **Questions and bugs** -- <https://github.com/miguelgandra/nautilus/issues>.
