---
title: "Reconstructing Movement Tracks"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Reconstructing Movement Tracks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

An animal-borne tag rarely knows where it is. Satellite fixes arrive only when the
animal surfaces -- minutes or hours apart, and never during a dive. Between those fixes,
though, the tag *does* know which way the animal is pointing and how fast it is moving. The
movement-tracks branch turns that into a continuous path: it **dead-reckons** a fine-scale
pseudo-track from heading, speed and depth, pulls the accumulated drift back onto the known
fixes, and -- most importantly -- lets you **measure how far to trust the result** before
you use it.

This guide covers that branch end to end. It assumes you already have a processed
`nautilus_tag`; if not, start with `vignette("getting-started", package = "nautilus")`,
which walks the mandatory pipeline up to `processTagData()`.

The code below is illustrative: it shows the real calls in order, but the chunks are not
executed here because reconstructing a track needs your own processed deployment.

## What you need, and the one thing to check first

Track reconstruction is a *pure integrator*. It consumes what `processTagData()` already
produced -- a per-sample `heading`, `pitch`, a swimming-speed channel, and `depth` -- and
turns them into positions. It does not re-estimate orientation, so the track is **only as
good as the heading that feeds it**.

That makes one upstream step non-negotiable: **calibrate the magnetometer first.** An
uncalibrated hard-iron offset is a roughly constant heading bias, and a constant heading
bias compounds into ever-growing positional drift. `reconstructTrack()` will warn loudly if
it sees an uncalibrated magnetometer, but the fix belongs upstream -- run
`calibrateMagnetometer()` (ideally with a dedicated calibration recording) and confirm the
status on a deployment (`tagMetadata(tags[[1]])$mag_calibration$status`) before reconstructing.

You also need **known positions** to anchor to. These live in the tag's canonical position
record: the deployment origin, any FastGPS/Argos fixes recorded at the surface, and the
pop-up location. `reconstructTrack()` reads them automatically.

## 1. Make sure the fixes are clean

Surface fixes are noisy. A single Fastloc fix with a bad fix geometry can sit tens of
kilometres off the real track, and if you anchor to it, the whole reconstruction bends
toward that error. Screening the fixes is therefore essential -- but it is a **data-quality
step, not a track step**, so it belongs in the clean / quality-control phase alongside the
sensor checks (see `vignette("getting-started")`). If you followed that pipeline, your
position record is already vetted and you can go straight to reconstruction.

To run or tune it here, `filterLocations()` screens the tag's GPS/Argos fixes and **removes**
any that imply an impossible swimming speed, were computed from too few satellites, or (as a
gross-error bound) sit implausibly far from the deployment. It uses the neighbour-consistency
"root" test of Freitas et al. (2008): a fix is a spike only when it implies an impossible
speed to *both* its neighbours, so a single genuinely fast segment is kept. Only the
automatically-acquired FastGPS/Argos fixes are removed -- user-entered positions and the
deploy/pop-up anchors are trusted and kept. All three checks are opt-in.

```{r filter-locations}
library(nautilus)

# `tags` is the processed object. Deployment coordinates for the (optional) distance check are
# taken from each tag's metadata; pass id.metadata = meta only to override or cross-check them.
tags <- filterLocations(
  tags,
  max.speed.kmh  = 10,     # reject fixes implying > 10 km/h to BOTH neighbours
  min.satellites = 4,      # drop Fastloc fixes computed from < 4 satellites
  plot           = FALSE   # set TRUE for a diagnostic map (coastline needs maps/mapdata)
)
```

Thresholds are species-specific: the 10 km/h above is a conservative cruising cap for a whale
shark. The distance check (`max.distance.km`) is off by default -- enable it only as a loose
bound on decoding errors, well beyond the animal's real range, since anchored to the
deployment it will otherwise clip genuine long-range displacement. The speed and distance
checks need the **geosphere** package.

## 2. Dead-reckon the track

`reconstructTrack()` integrates heading and speed forward, step by step, on a spherical
Earth, attaches the measured depth as the vertical axis, and then applies **Verified
Position Correction (VPC)** to redistribute the reckoning drift so the path lands on each
known fix. All of its behaviour is grouped in `reconstructTrackControl()`.

```{r reconstruct}
tracks <- reconstructTrack(
  tags,
  control = reconstructTrackControl(
    speed.method = "paddle",         # where swimming speed comes from
    vpc.method   = "error_weighted"  # how drift is reconciled with the fixes
  )
)
```

The result is a named list, one element per deployment, with `pseudo_lon`, `pseudo_lat`,
`pseudo_depth`, `speed_dr` (the speed actually used) and `pseudo_error` (a per-sample
1-sigma positional uncertainty, in metres) added to each object.

**Choosing a speed method.** Heading sets the *shape* of the track; speed sets its *scale*.
`reconstructTrackControl(speed.method = ...)` offers four sources:

- `"paddle"` -- the measured paddle-wheel speed. The natural choice for PILOT tags that
  carry a paddle wheel.
- `"vedba"` -- speed inferred from a VeDBA activity model, auto-calibrated from the
  deployment's own GPS fixes when `vedba.model = NULL`.
- `"depth_rate"` -- speed from dive geometry (vertical velocity and pitch); trusted only on
  steep segments (`|pitch| >= depth.rate.min.pitch`, default 45 degrees) and back-filled
  with a constant elsewhere.
- `"constant"` -- a single fixed speed (`constant.speed`, default 0.5 m/s). Produces a
  *shape-faithful* pseudo-track when no speed sensor is trustworthy, and is the fallback
  whenever any other method returns a non-finite value.

**Choosing a VPC method.** `"error_weighted"` (the default) reconciles the reckoned path
with each fix using a Kalman-style gain based on the fix's own error radius, so a precise
FastGPS fix pulls the track hard while a coarse Argos fix barely nudges it. `"linear"`
forces the path exactly through every fix. `"scale_rotate"` applies a single scale-and-rotate
per segment -- best when the drift is *systematic* (a mis-scaled speed or a constant heading
bias). `"none"` returns the uncorrected reckoning.

**The reconstructability gate.** For deployments with *no interior fixes* -- anchored only by
the deployment and pop-up -- the shape of the track is essentially unconstrained, so
`reconstructTrack()` reports a soft reliability check. It computes *directedness* (net
start-to-end displacement divided by the reckoned path length) and, when it falls below
`reconstructability.min` (default 0.1), warns that the reconstruction is weakly supported.
It never aborts -- a directed track is still returned -- but treat a flagged track with
caution. This is a triage heuristic, not a verdict; where you *do* have interior fixes, the
next step measures accuracy directly.

## 3. Validate before you trust it

A pseudo-track always looks plausible -- it is a smooth line through your fixes. The question
is how far it strays from the truth *between* them, and the honest answer comes from
`crossValidateTrack()`. It performs leave-one-out cross-validation: it withholds each genuine
fix in turn, reconstructs the track from the remaining anchors (so neither the VPC nor the
VeDBA calibration ever sees the withheld fix), and measures the distance from the
reconstructed position to the true one.

```{r cross-validate}
cv <- crossValidateTrack(
  tags,                                # the processed object, NOT the reconstructed track
  control = reconstructTrackControl(speed.method = "paddle", vpc.method = "error_weighted")
)
```

Pass `crossValidateTrack()` the **processed object** and the **same control settings** whose
accuracy you want to assess -- it re-runs the reconstruction internally. The result is one
row per withheld fix:

| column | meaning |
|---|---|
| `error_m` | great-circle distance between reconstructed and true position |
| `gap_h` | hours to the nearest *retained* fix |
| `interpolated` | `TRUE` if retained fixes bracket the withheld one, `FALSE` if extrapolated |
| `fix_radius_m` | the withheld fix's own measurement radius (a floor on achievable error) |
| `speed_method`, `vpc_method` | the settings used, so you can stack and compare runs |

The most useful view is `error_m` against `gap_h`. If error grows roughly linearly with the
gap, the slope is an **empirical drift rate in metres per hour** -- divide it by 3600 to get
a `drift.rate` (m/s) you can feed back into `reconstructTrackControl()` so the reported
`pseudo_error` reflects your deployment's real drift. `crossValidateTrack()` prints the
median and 90th-percentile error and that empirical drift in its summary, and `plot = TRUE`
draws the error-versus-gap scatter and an error distribution.

One honest caveat, stated plainly in the function's own documentation: a held-out error
**conflates** heading error, speed error and unmodelled current. It tells you the net
accuracy of the whole pipeline, not which component failed. To isolate the heading itself,
compare the magnetometer against the geomagnetic model and the gyroscope, and use the next
tool to spot-check orientation against video.

### Sanity-checking the heading against video

If your tag carried a camera, you can check the heading that drives the track at the moments
it matters most. `findValidationSegments()` scans the processed data for the clearest
manoeuvres -- the sharpest turns, strongest rolls and steepest dives -- and returns short
time windows to inspect against the synchronised footage.

```{r validation-segments}
segs <- findValidationSegments(tags, types = c("turn", "roll", "dive"), n = 5)
# each row is a [start, end] window around a peak manoeuvre; open the video there
```

It returns the top `n` non-overlapping windows per manoeuvre type per animal, each with a
`peak_time`, an interpretable `value` and its `unit`. This validates the *orientation* (does
the computed turn match the observed one?), complementing the *positional* accuracy that
`crossValidateTrack()` measures.

## 4. Summarise the path

Once you trust a track, `trackMetrics()` reduces each one to a single row of trajectory
metrics -- path length, net displacement, and a family of tortuosity and straightness
indices that describe how convoluted the path is (a common proxy for behavioural mode:
directed travel versus area-restricted search).

```{r track-metrics}
metrics <- trackMetrics(
  tracks,                                              # the reconstructed track
  control = trackMetricsControl(metrics = "all", min.points = 5)
)
```

With `lon.col`/`lat.col` left `NULL`, `trackMetrics()` automatically uses `reconstructTrack()`'s
`pseudo_lon`/`pseudo_lat`. It returns `Total_distance_km` and `Net_displacement_km` alongside
`Path_ratio`, `Straightness` (net / total), `Sinuosity` (the Bovet & Benhamou index),
`Fractal_dimension`, `Mean_turning_angle`, and rolling `Hourly_tortuosity` /
`Daily_tortuosity` (window sizes set on the control). Metrics a track is too short to support
are returned as `NA` rather than guessed.

## 5. Map it

`plotTracks()` draws a per-animal grid overlaying the surface fixes, the deployment and pop-up
anchors, and the reconstructed pseudo-track.

```{r plot-tracks}
plotTracks(tracks, color.by = "depth")
```

FastGPS/Argos/user fixes, the deploy and pop-up anchors, and the pseudo-track are drawn in
distinct styles, with the track's start and end marked. Crucially, the track carries its
**uncertainty**: with `show.uncertainty = TRUE` (the default) a translucent corridor is drawn
from `pseudo_error`, so a segment that passes close to a fix reads tighter than one that has
drifted mid-gap; and `color.by = "depth"` (or `"speed"`) colours the track by the reconstructed
depth or speed on a shared scale. The land outline is a lightweight optional layer (needs
**maps**/**mapdata**, silently skipped if absent -- no tile server or network); set
`show.bathymetry = TRUE` to overlay depth contours via **marmap**.

## 6. Hand off to a state-space model

A pseudo-track is a reconstruction, not a set of georeferenced observations. For analyses
that need a formally smoothed track with per-position credible intervals -- utilisation
distributions, behavioural-state estimation -- delegate to a continuous-time state-space
model. `nautilus` does not re-implement an SSM smoother; it formats the track for the
well-tested packages that do.

```{r export-ssm}
ssm_input <- exportForSSM(tracks, thin.minutes = 10)
# columns: id / date / lc / lon / lat / x.sd / y.sd
```

`exportForSSM()` thins the dense track to one position per `thin.minutes` and passes the
reckoning uncertainty (`pseudo_error`) through as per-point observation error in the `x.sd` /
`y.sd` columns (in metres) that **aniMotum** (>= 1.2) expects. With aniMotum installed:

```{r animotum, eval = FALSE}
fit <- aniMotum::fit_ssm(ssm_input, model = "crw", time.step = 2)  # 2-hour steps
aniMotum::grab(fit, "predicted")
```

crawl users can project the coordinates and pass the metre-scale `y.sd` to `crwMLE()`'s error
model.

## Where to go next

- **Getting Started** -- the mandatory pipeline that produces the processed tag this guide
  starts from: `vignette("getting-started", package = "nautilus")`.
- **Function reference** -- `?reconstructTrack`, `?reconstructTrackControl`,
  `?crossValidateTrack`, `?trackMetrics`, `?exportForSSM`.
- **Calibration** -- because the track is only as good as its heading:
  `?calibrateMagnetometer`.
- **Questions and bugs** -- <https://github.com/miguelgandra/nautilus/issues>.
