---
title: "Orientation methods: what nautilus corrects, and what it cannot"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Orientation methods: what nautilus corrects, and what it cannot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

The function documentation for `processTagData()` and `checkSensorIntegrity()` states the
practical limits of nautilus orientation estimates in a sentence or two. This vignette is the
long version: it sets out the frames of reference involved, derives how far each uncorrected
error propagates into pitch, roll, heading and dead-reckoned speed, and explains why some of
those errors are invisible to the package's automated checks.

It is aimed at readers deciding how much to trust a derived orientation series, or diagnosing a
deployment whose kinematics look wrong. Every number below is reproduced by the code shown; the
chunks are not evaluated when the vignette is built, so you can copy them into a session and
check the results yourself.

## 1. Three frames of reference

A tag records in its **sensor frame**: whatever the accelerometer, gyroscope and magnetometer
chips are physically aligned with inside the housing. Biology happens in the **animal frame**:
`x` forward along the body axis, `y` out to one side, `z` up through the back. Between them sits
the **package frame**, the orientation of the housing on the animal, which is fixed for a given
deployment but differs between deployments and between attachment methods.

nautilus resolves the difference in two separate steps, and the distinction matters because they
have very different reliability.

**Step 1, the discrete part.** `checkTagMapping()` and `applyAxisMapping()` resolve which sensor
axis corresponds to which body axis, and in which direction. This is a search over the 24 proper
rotations that map one right-handed axis triple onto another - six axis permutations times four
sign combinations that preserve handedness. It is a *discrete* choice, so it is either right or wrong, with no partial credit.
It is not, however, guaranteed to resolve: the vertical axis is resolved from gravity and is
reported "gravity-ambiguous" when mount tilt obscures it, while surge versus sway is resolved from
*diving* signal (the pitch-versus-depth-rate relationship) and is reported "ambiguous" when the
animal did not dive enough. That is precisely why `consensusAxisMapping()` (pool the decision over a
package) and `reviewTagMapping()` (confirm it by eye) exist - use them rather than assuming a
mapping was resolved.

**Step 2, the continuous part.** Once the axes are named correctly, the housing is still not
perfectly aligned with the animal: it sits at some small pitch and roll relative to the body axis,
and it may be rotated about the animal's vertical axis. These are continuous offsets, they must be
*estimated* rather than selected, and this is where the limits are.

## 2. Pitch and roll mounting offsets

`processTagData()` estimates a constant pitch and roll offset for the deployment and subtracts it
(controlled by `correct.pitch` and `correct.roll` in `orientationControl()`). The two offsets use
different estimators, and knowing which is which explains what they can and cannot see.

**Pitch** uses the Kawatsu method: regress pitch (in radians) on the smoothed vertical velocity
and take the **intercept** - that is, the pitch the tag reports when the animal is not going up or
down. A correctly-mounted tag reads zero there; a nose-down mounting reads its own offset. The
correction is applied only when the regression is strong enough (`pitch.offset.min.r2`) and the
resulting offset is physically plausible (`warning.threshold`), because a weak fit
means the "offset" is really just the animal's mean pitch, and subtracting that would strip
genuine posture signal.

**Roll** uses the median roll over the most level half of the record (the samples whose *absolute*
smoothed vertical velocity is below its own median). A bilaterally symmetric animal cruises upright on
average, so any persistent roll during level swimming is attributed to the mounting.

Both assume the animal spends a substantial part of the record close to level, and both estimate a
single constant shift. A rigid shift can remove a constant bias but cannot remove an error that
*scales with the animal's own attitude*. The next section is about exactly such an error - and
about the fact that both estimators are blind to it by construction.

## 3. Mounting yaw: the rotation that is not estimated

If the tag is rotated by an angle `psi` about the animal's vertical axis - nose of the tag not
pointing along the nose of the animal - nautilus does not estimate or remove it. This is stated
in the `processTagData()` documentation; here is what it costs.

Write the gravity vector in the animal frame for a given pitch and roll, then re-express it in a
tag frame yawed by `psi`, and compute pitch and roll from it the way a tilt-compensated compass
does:

```{r frames}
d2r <- pi / 180; r2d <- 180 / pi

grav_animal <- function(pitch, roll) {
  p <- pitch * d2r; r <- roll * d2r
  c(-sin(p), sin(r) * cos(p), cos(r) * cos(p))
}
yawR <- function(psi) {
  s <- sin(psi * d2r); c_ <- cos(psi * d2r)
  matrix(c(c_, s, 0, -s, c_, 0, 0, 0, 1), 3, 3, byrow = TRUE)
}
pr_from_g <- function(g) {
  c(pitch = -asin(max(-1, min(1, g[1]))) * r2d,
    roll  = atan2(g[2], g[3]) * r2d)
}
```

### 3.1 Pitch is compressed, with no offset

The exact relation is `sin(observed_pitch) = cos(psi) * sin(true_pitch)`: a yawed tag reads a
*compressed* version of the animal's true pitch, by a factor that approaches `cos(psi)` at small
angles. Regressing observed on true pitch over a realistic range:

```{r pitch-scale}
psi <- 30
true_p <- seq(-60, 60, by = 0.5)
obs <- t(sapply(true_p, function(p) pr_from_g(as.vector(yawR(psi) %*% grav_animal(p, 0)))))
signif(coef(lm(obs[, "pitch"] ~ true_p)), 6)
#>  (Intercept)       true_p
#> -4.25144e-16  8.35277e-01
```

The slope is 0.8353 (against `cos(30 deg) = 0.8660`; the relation is not exactly linear over
+/-60 degrees, so the least-squares slope sits slightly below the small-angle limit). **The
intercept is zero** - the printed -4.25e-16 is floating-point noise. Mounting yaw is a pure *scale* error on pitch, with no offset
whatsoever - and the Kawatsu estimator of section 2 uses only the intercept. It is therefore blind
to yaw by construction, and no amount of extra data will reveal it.

### 3.2 A phantom roll appears on a level animal

The same rotation projects part of the animal's pitch onto the tag's roll axis, so an animal that
is not rolled at all appears to be banking whenever it pitches:

```{r phantom-roll}
for (p in c(30, 45)) print(pr_from_g(as.vector(yawR(30) %*% grav_animal(p, 0))))
#>  pitch   roll
#>   25.7   16.1
#>  pitch   roll
#>   37.8   26.6
```

At a 30 degree descent the tag reports 25.7 degrees of pitch and **16.1 degrees of roll on an
animal with zero roll**. At 45 degrees it reports 26.6 degrees of phantom roll. If you are
scoring rolling behaviour, banking turns or prey-capture manoeuvres from roll, a mounting yaw
manufactures those events out of ordinary diving.

### 3.3 Dead-reckoned speed inherits the error

`reconstructTrack(speed.method = "depth_rate")` uses the orthogonally-corrected depth rate,
`speed = |vertical velocity| / sin(|pitch|)`. Because pitch is compressed, the denominator is too
small and speed is over-estimated:

```{r ocdr}
p_true <- 45
p_obs  <- pr_from_g(as.vector(yawR(30) %*% grav_animal(p_true, 0)))["pitch"]
100 * (sin(p_true * d2r) / sin(abs(p_obs) * d2r) - 1)
#> 15.5
```

A 30 degree mounting yaw inflates depth-rate speed by 15.5%. Note that this bias is the *same at
every dive angle*: from `sin(observed) = cos(psi) * sin(true)` the ratio is exactly `1/cos(psi)`,
independent of pitch. It is a clean multiplicative bias on speed, so it propagates undiluted into
total distance travelled and into every downstream track metric.

### 3.4 Both mounting-offset estimators pass it clean

The two previous sections are geometry. This one runs nautilus's *actual* estimators over a
simulated record - 200 v-shaped dives to +/-40 degrees, true roll approximately zero, 1 m/s - and
asks what they report.

```{r estimators}
set.seed(7)
n <- 120000                                   # 200 v-shaped dives
true_pitch <- 40 * sin(seq(0, 200 * 2 * pi, length.out = n)) + rnorm(n, 0, 3)
true_roll  <- rnorm(n, 0, 4)                  # the animal's REAL roll: noise about zero, SD 4 deg
vv         <- 1 * sin(true_pitch * d2r)       # vertical velocity at 1 m/s

estimators <- function(psi, pitch_offset = 0) {
  p <- true_pitch + pitch_offset
  G <- t(mapply(function(a, b) as.vector(yawR(psi) %*% grav_animal(a, b)), p, true_roll))
  o <- t(apply(G, 1, pr_from_g))
  vvs <- as.numeric(stats::filter(vv, rep(1 / 201, 201), sides = 2))   # smoothed vertical velocity
  ok  <- is.finite(vvs)
  kaw <- coef(lm(I(o[ok, "pitch"] * d2r) ~ vvs[ok]))[1] * r2d          # Kawatsu: the intercept
  lev <- ok & abs(vvs) <= median(abs(vvs), na.rm = TRUE)               # the "level half"
  c(pitch_intercept = unname(kaw),
    roll_median     = median(o[lev, "roll"], na.rm = TRUE),
    roll_sd_dives   = sd(o[!lev, "roll"], na.rm = TRUE))
}
round(rbind(`no yaw`             = estimators(0),
            `30 deg yaw`         = estimators(30),
            `30 deg yaw + 8 deg` = estimators(30, 8)), 3)
#>                    pitch_intercept roll_median roll_sd_dives
#> no yaw                      -0.001      -0.013         4.009
#> 30 deg yaw                   0.003       0.054        20.502
#> 30 deg yaw + 8 deg           6.665       4.141        20.878
```

Read the middle row carefully. Both estimators report essentially zero - **there is nothing for
them to correct, and they are right by their own criteria** - while the apparent rolling of the
animal has been inflated fivefold, from 4.0 to 20.5 degrees of standard deviation, entirely by the
mounting. The deployment passes every mounting-offset diagnostic nautilus has and still yields a
roll series that is mostly artefact.

The third row adds a further consequence. When a genuine +8 degree mounting pitch *is* present
alongside the yaw, Kawatsu recovers only 6.67 degrees of it - short by close to the `cos(psi)` factor
(6.665/8 = 0.833; the exact geometric prediction is `asin(cos(30 deg) * sin(8 deg)) = 6.92 deg`, a
ratio of 0.865, with the remainder down to the least-squares fit discussed in 3.1). Mounting yaw
does not merely go undetected: it also biases the corrections that are applied, and it manufactures a 4.14 degree roll offset that the roll corrector then dutifully
subtracts from a record whose real roll offset is zero.

### 3.5 What to do about it

There is no automated remedy in nautilus 0.1.0. The practical mitigations are:

- **Attach carefully.** Yaw is the one mounting error that instrumentation discipline can remove
  outright; a few degrees costs almost nothing (the compression goes as `cos(psi)`, so 5 degrees
  costs 0.4%), while 30 degrees costs 15%.
- **Record the attachment geometry.** A photograph at deployment is usually enough to bound `psi`.
- **Cross-check with camera footage** where available (`renderOverlayVideo()`), which will show a
  persistent roll offset during straight-line swimming.
- **Treat roll cautiously** on any deployment where roll correlates with pitch. That correlation
  is the signature of mounting yaw and should not be interpreted as banking behaviour.

## 4. Per-axis gain errors and the limits of `accel.scale`

`checkSensorIntegrity(checks = "accel.scale")` tests whether the static acceleration magnitude is
near 1 g. It is a scalar test on a vector quantity, so it is powerful against errors that affect
all three axes and weak against errors that affect one.

Simulate a 20% gain error on `ay` alone, for an animal swimming close to level:

```{r per-axis}
set.seed(42)
n <- 2e5
p_t <- rnorm(n, 0, 12); r_t <- rnorm(n, 0, 18)
G  <- t(mapply(function(p, r) grav_animal(p, r), p_t, r_t))
Gb <- G; Gb[, 2] <- Gb[, 2] * 1.20

median(sqrt(rowSums(Gb^2)))                       # what accel.scale measures
#> 1.0093

sd(atan2(G[, 2],  G[, 3]) * r2d)                  # true roll SD
#> 17.98
sd(atan2(Gb[, 2], Gb[, 3]) * r2d)                 # roll SD with the faulty axis
#> 20.81
```

The static magnitude reads 1.0093 g. The default tolerance is `accel.scale.tol = 0.2`, so this
fault is **not flagged** - and it would not be flagged at 0.1, or 0.05, or 0.02 either. Yet the
roll series is visibly wrong: its standard deviation is inflated from 18.0 to 20.8 degrees, with
a mean absolute error of 2.4 degrees and a maximum of 5.2 degrees.

The reason is geometric. On a near-level animal, `ay` is small, so scaling it barely moves the
vector magnitude; but roll is `atan2(ay, az)`, which depends on `ay` directly. The quantity the
check measures is precisely the quantity the fault does not perturb.

This is why the `checkSensorIntegrity()` documentation says the check "may not detect errors
affecting only individual axes", and why it advises inspecting individual sensor axes as well as
the summary metric. Note that `plot = TRUE` will not help here: the report draws a page only for
deployments that raised an error or warning, with one panel per *firing check* - and the whole point
of this fault is that nothing fires. Inspecting it means plotting the axes yourself, or comparing
roll against a known-good deployment.

### 4.1 Why the tolerance is 0.2

The default was set from 54 real deployments spanning two tag makes (CATS and Little Leonardo):

| cohort | n | median static \|a\| | max \|median - 1\| |
|---|---|---|---|
| CATS, whale shark | 51 healthy | 0.895 - 1.030 | 0.105 |
| CATS, reef manta | 1 | 0.962 | 0.038 |
| Little Leonardo, hammerhead | 1 | 0.982 | 0.018 |
| CATS, one faulty deployment | 1 | 4.095 | 3.095 |

The worst healthy deployment deviates by 0.105, so a tolerance of 0.2 leaves roughly a factor of
two of headroom and produced zero false positives across the cohort. The faults it is meant to
catch are far outside it: acceleration left in m/s^2 reads about 9.4 (deviation 8.4), a doubled
gain reads 2.0, a halved gain 0.5. Tightening to 0.10 would have flagged two apparently healthy
deployments; tightening to 0.05 would have flagged ten.

The single flagged deployment was genuine: one accelerometer axis was pinned at the +4 g range
limit for 98.7% of an 11.4-million-sample record.

## 5. Gyroscope bias

`checkSensorIntegrity(checks = "gyro.bias")` reports a persistent per-axis gyroscope offset.
Gyroscope bias is **not** explicitly corrected anywhere in nautilus, and what that costs depends
entirely on which orientation algorithm you use:

- **`orientation.algorithm = "tilt_compass"` (the default)** is a 6-axis method using only the
  accelerometer and magnetometer. It never reads the gyroscope, so gyroscope bias cannot affect
  it at all. For the large majority of users this check is purely informational.
- **`orientation.algorithm = "madgwick"`** integrates the gyroscope, so bias does enter. The
  filter's accelerometer and magnetometer correction term continuously pulls the estimate back
  toward the gravity and magnetic references, which absorbs small biases; the implementation does
  not, however, include an explicit bias-estimation state, so it relies on that feedback alone.
  Larger biases will degrade the orientation estimate and should be inspected before you rely on
  derived movement metrics.

If a deployment reports a substantial gyroscope bias and you need `madgwick` (for example because
the magnetometer is contaminated by a paddle wheel), compare the two algorithms on the same record
before deciding which to trust.

## 6. Summary

| error | estimated by nautilus? | detectable by the automated checks? | main consequence |
|---|---|---|---|
| axis permutation / sign | yes, `applyAxisMapping()` | yes, but it may report "ambiguous" rather than resolve | catastrophic if wrong, so it is checked hard |
| pitch / roll mounting offset | yes, when the animal is often level | partly, via the fit gates | constant attitude bias |
| mounting yaw | **no** | **no** - both estimators return zero | compressed pitch, phantom roll, inflated OCDR speed, and biased recovery of the offsets that *are* corrected |
| uniform accelerometer scale error | no | yes, `accel.scale` | wrong units propagate everywhere |
| single-axis gain error | no | **no** - invisible to a scalar magnitude test | inflated roll, subtly wrong kinematics |
| gyroscope bias | no | yes, `gyro.bias` (informational) | none under `tilt_compass`; drift under `madgwick` |

The recurring theme is that nautilus is good at the errors that change a vector's *length* and
blind to several that change only its *direction*. Where an automated check cannot see a problem,
the documentation says so rather than implying a guarantee it cannot make.

## See also

- `processTagData()` - orientation estimation and the mounting-offset controls
- `orientationControl()` - `correct.pitch`, `correct.roll`, `madgwick.beta`
- `checkSensorIntegrity()` - `accel.scale`, `gyro.bias` and the other channel checks
- `checkTagMapping()`, `reviewTagMapping()`, `applyAxisMapping()` - the discrete axis resolution
- `vignette("movement-tracks", package = "nautilus")` - how speed and heading enter the track
